
```{r setup, echo=FALSE, warning=FALSE, purl=FALSE, message=FALSE}
options(repos = "http://cran.rstudio.com/")
pkgs <- c("dplyr", "knitr")
x<-lapply(pkgs, library, character.only = TRUE)
opts_chunk$set(tidy = FALSE, message = F, warning = F)
```

# Understanding metabolism

Get the lesson R script: [understanding.R](understanding.R)

Get the lesson data: [download zip](data/779980.zip)

## Lesson Outline

- [Goals and motivation]
- [Summarizing and plotting metabolism data]
- [Comparing metabolism with water quality data]
- [Comparing metabolism with nutrient data]

## Lesson Exercises

- [Exercise 1]
- [Exercise 2]
- [Exercise 3]

## Goals and Motivation

This lesson will teach you how to begin evaluating estuary metabolism estimates using additional data available in SWMP. Our goals are to understand the tools that are available in R, and the tidyverse in particular, to combine and explore relationships among variables that could influence metabolism. These tools are the building blocks of more detailed analyses that will help you understand site-specific relationships or to develop more comprehensive comparisons between reserves.  

By the end of this lesson, you should know how or find the resources to be able to do the following:

* Summarize metabolism data using dplyr
* Make some basic plots of metabolism using ggplot2
* Combine metabolism estimates with additional SWMP data using dplyr
* Compare metabolism estimates with additional SWMP data using ggplot2

## Summarizing and plotting metabolism data

The [SWMPr](https://cran.r-project.org/web/packages/SWMPr/index.html) and [WtRegDO](https://github.com/fawda123/wtregdo) packages provide a set of easy to use functions for working with SWMP data to estimate metabolism.  Our goals so far have been to demonstrate how to use the pre-existing functions to evaluate metabolism in our example datasets.  

What we have not covered is how to use the other tools available in R to create some more customized examples of data analyses.  This is obviously a huge topic that we cannot adequately cover in 90 minutes.  However, our goal today is to provide you with some general building blocks for developing some more specific analyses. We'll be leveraging some of the other packages in the tidyverse to accomplish this goal.

We start with our detided dissolved oxygen data from Apalachicola. Check out the [detiding](detiding.html) to review to see how we got these data. For sake of time, you can import this data file from the website: 

```{r}
load(url('https://tbep-tech.github.io/ecometab-r-training/data/apadtd.RData'))
head(apadtd)
```

For this exercise, we'll assume the detided data is appropriate for estimating metabolism.  Remember, in practice you'll want to check out the data with `evalcor` and evaluate the metabolism results with `meteval`.

We estimate metabolism as follows, using the detided dissolved oxygen data:

```{r}
tz <- attr(apadtd$DateTimeStamp, which = 'tzone')
lat <- 29.6747
long <- -85.0583

apaeco <- WtRegDO::ecometab(apadtd, DO_var = "DO_nrm", tz = tz, lat = lat, long = long)
head(apaeco)
```

The WtRegDO package has a plotting function that lets you view the metabolism results. This function lets you view the daily estimates or as aggregated summaries of the daily results.  For follow-up analysis, it will be useful to understand how to plot the data using more general plotting functions.  This will allow you to conduct your own analyses outside of the "pre-packaged" summaries in WtRegDO. 

For plotting, we can use the ggplot2 library.  This comes with the tidyverse, so no need to install it separately, but you'll of course need to load the library. 

```{r}
library(ggplot2)
```

With ggplot2, you begin a plot with the function `ggplot()`. This creates a coordinate system that you can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = apaeco)` creates an empty base graph.

```{r, eval = F}
ggplot(data = apaeco)
```

The next step is to "map" variables in your dataset to the plot visual properties or "aesthetics" (i.e., x variable, y variable, etc.).   The `mapping` argument is defined with `aes()`, and the `x` and `y` arguments of `aes()` specify which variables to map to the x and y axes. The functoin looks for the mapped variable in the `data` argument, in this case, `apaeco`.  

```{r, eval = F}
ggplot(data = apaeco, aes(x = Date, y = Pg))
```

Now we need to add a "geometry" or `geom` to the plot to add components for viewing.  We can add one or more layers (aka `geoms`) to `ggplot()`. The function `geom_point()` adds a layer of points to your plot based on the variable mappings in the `aes` function.

```{r, fig.height = 3, fig.width = 7}
ggplot(data = apaeco, aes(x = Date, y = Pg)) + 
  geom_point()
```

We can also overlay a line using `geom_line()`. 

```{r, fig.height = 3, fig.width = 7}
ggplot(data = apaeco, aes(x = Date, y = Pg)) + 
  geom_point() + 
  geom_line()
```

Those are the basics of ggplot.  This structure may seem weird at first, but it's design is purposeful and easily extendable.  Just remember these requirements:

* All ggplot plots start with the `ggplot` function
* It will need three pieces of information: the **data**, how the data are **mapped** to the plot **aesthetics**, and a **geom** layer

The daily data are a bit messy, so maybe we want to aggregate to averages before plotting to make the trends a little more apparent.  We need to use a few functions from the lubridate and dplyr packages to do this.  As before, these come with the tidyverse package. 

The lubridate package provides a set of functions to quickly modify date and time objects in R.  It provides many functions to convert R vectors to date objects (e.g., a character strings of dates imported from a csv) and many functions to extract components of date objects (e.g., month, day, etc.).  The latter is extremely useful for analysis.  

First we load the lubridate and dplyr packages.  Then we use both to create a month column from our existing date column in the apaeco data set. The `mutate` function from dplyr adds a column to our dataset and the `month` function extracts the month from the existing date column. 

```{r}
library(lubridate)
library(dplyr)

apasum <- mutate(apaeco, mo = month(Date))
head(apasum)
```

We can plot the monthly values as boxplots with ggplot using the `geom_boxplot` geometry. 

```{r, fig.height = 3, fig.width = 7}
ggplot(apasum, aes(x = mo, y = Pg, group = mo)) + 
  geom_boxplot()
```

Or, if we want to summarize the raw data, we can use the dplyr `group_by()` and `summarise()` functions.  Here we are grouping the data by month, then taking the average production values. Note the use of `na.rm = T` in the `mean` function. 

```{r}
apasum <- group_by(apasum, mo)
apasum <- summarise(apasum, meanpg = mean(Pg, na.rm = T))
```

Then we plot as before using `geom_point()` and `geom_line()`.

```{r, fig.height = 3, fig.width = 7}
ggplot(apasum, aes(x = mo, y = meanpg)) + 
  geom_point() +
  geom_line()
```

## Exercise 1

Now repeat some of the above analyses using the Sapelo Island metabolism data. 

1. Load the detided Sapelo Island data (`sapdtd`) using this code: 
    ```{r, eval = F}
    load(url('https://tbep-tech.github.io/ecometab-r-training/data/sapdtd.RData'))
    ```
1. Estimate metabolism using the `ecometab` function from the WtRegDO package. Run `WtRegDO::ecometab` using `tz = 'America/Jamaica'`, `lat = 31.39`, and `long = -81.28`.
1. Create a month variable for `sapdtd` using the `mutate` and `month` functions.
1. Group the data by month and summarize the average production using `group_by` and `summarise`. 
1. Plot the results with `ggplot` by mapping the month and mean production columns to the x, y aesthetics and then using `geom_point()` or `geom_line()`

## Comparing metabolism with water quality data

One of the values of SWMP data is that we get a lot of supporting information that we can use to describe water quality changes.  For metabolism, we might expect that it tracks physical drivers like temperature or light energy, or perhaps nutrient inputs.  We can combine the metabolism estimates with other sWMP parameters to evaluate these hypotheses. 

We can load SWMP data as before using tools from SWMPr, in particular using the `import_local`, `qaqc`, and `comb` functions.  We've already done this a few times for the dry bar station at Apalachicola (see the "data prep" section in the [metabolism](odum.html) lesson). Our detided dataset from above includes some relevant water quality and weather data that we can evaluate.  

For example, perhaps we want to compare water temperature to metabolism. Since our metabolism estimates are on the daily scale, we'll also want to summarize our water quality data at this scale. We can use tools in the lubridate and dplyr packages to do this, just as above. 

We can add a date column, group by day, then take average water temperature. 

```{r}
# add a date column
apasum <- mutate(apadtd, Date = date(DateTimeStamp))

# group by date
apasum <- group_by(apasum, Date)

# summmarise average water temperature
apasum <- summarise(apasum, meantemp = mean(Temp, na.rm = T))
```

The last thing to do is to join these summarized data with the metabolism results. For this, we can use a "join" function from dplyr.  We tell the `inner_join` function to combine the datasets using the `Date` column that is included in both the summarized water quality data and metabolism results. 

```{r}
# join water quality with metabolism
apasum <- inner_join(apasum, apaeco, by = 'Date')
```

Now we're ready to plot.  We can make a quick scatterplot with ggplot by mapping the appropriate variables in our dataset to the x, y aesthetics. Temperature is our independent variable in this case, so we put it on the x-axis.  

```{r, fig.height = 5, fig.width = 5}
p <- ggplot(data = apasum, aes(x = meantemp, y = Pg)) + 
  geom_point()
p
```

Not much there, so maybe we can want to try exploring some other relationships. 

## Exercise 2

Repeat the above examples using the Sapelo Island dataset.

1. First, estimate metabolism for the Sapelo Isalnd data.  You should have this already from the first exercise. Here it is again:
    ```{r, eval = F}
    load(url('https://tbep-tech.github.io/ecometab-r-training/data/sapdtd.RData'))
    sapeco <- WtRegDO::ecometab(sapdtd, DO_var = 'DO_nrm', lat = 31.39, long = -81.28, tz = 'America/Jamaica')
    ```
1. From the detided Sapelo Island dataset, create a date column with the `mutate` and `date` functions.  
1. Summarize mean daily water temperature using `group_by` and `summarise`. 
1. Join the summarized data with the metabolism data using `inner_join` and `by = 'Date'`. 
1. Make a scatterplot in ggplot by mapping the mean temperature and production summaries to the x, y aesthetics. 

## Comparing metabolism with nutrient data

Our last example showed that we couldn't find a very strong relationship of water temperature with production.  That's okay.  We had a hypothesis, checked hte data, and then made a general conclusion that temperature isn't a strong driver of production at the Apalachicola dry bar site.  This doesn't mean temperature isn't always important - the limited data we have and our analysis at the daily scale provides us with some clues, but there's much more we can evaluate. 

Now we have the tools to start looking at some additional datasets.  Another reasonable expectation is that metabolism may be driven by nutrients.  We can us the same tools above to import, combine, and evaluate relationships of metabolism with nutrients. 

Let's import nutrient data from the dry bar station using `import_local` from SWMPr. 

```{r}
apadbnut <- import_local(path = 'data', station_code = 'apadbnut')
```

We can use the `qaqc` function as before to handle the QA/QC flags. 

```{r}
apadbnut <- qaqc(apadbnut, qaqc_keep = c('0', '1', '2', '3', '4', '5'))
head(apadbnut)
```

If we want to compare any of the nutrient data to metabolism, we need to aggregate the information at a time scale that allows comparison.  The metabolism results are daily, but the nutrient data are collected monthly.  

We can first try joining the nutrient data to metabolism by day.  First we need to create a date column for the nutrients.

```{r}
apadbnut <- mutate(apadbnut, Date = date(datetimestamp))
```

Now we can join by date.  We use `inner_join` to combine by dates that are shared between the two datasets. 

```{r, fig.height = 5, fig.width = 5}
apajoin <- inner_join(apaeco, apadbnut, by = 'Date')
```

We can now plot the metabolism estimates against the various nutrient observations. Here's an example looking at chlorophyll.

```{r, fig.height = 5, fig.width = 5}
ggplot(apajoin, aes(x = chla_n, y = Pg)) + 
  geom_point()
```

Again, not super interesting. As a final attempt at comparing these data, we can try averaging results to a common unit of time, such as month.  This may reduce some of the variability in the data and also provide a more general comparison by evaluating average conditions as opposed to "instantaneous" comparisons at the daily scale.  

For both the metabolism and nutrient data, we can "floor" the dates to the first of each month, group by date, then summarize average values.  Although technically we're changing the dates, we are retaining the date class in the date column which can sometimes help with plotting or other downstream analysis.  We use the `floor_date` function from lubridate. 

```{r}
# floor nutrient dates
apadbnut <- mutate(apadbnut,
    dateflr = floor_date(Date, unit = 'month')
  )

# floor metabolism dates
apaeco <- mutate(apaeco, 
    dateflr = floor_date(Date, unit = 'month')
  )
```

Then group by the floored dates, summarize to averages, and join the data.

```{r}
# group by and summarize apadbnut
apadbnut <- group_by(apadbnut, dateflr)
apadbnut <- summarise(apadbnut, 
  chla_n = mean(chla_n, na.rm = T), 
  nh4f = mean(nh4f, na.rm = T)
  )

# group by and summarize metabolism
apaeco <- group_by(apaeco, dateflr)
apaeco <- summarise(apaeco, 
  Pg = mean(Pg, na.rm = T), 
  Rt = mean(Rt, na.rm = T), 
  NEM = mean(NEM, na.rm = T)
  )

# join
apajoin <- inner_join(apaeco, apadbnut, by = 'dateflr')
```

Now plot as before.  Here we're looking at average Production vs chlorophyll using the monthly summaries.

```{r, fig.height = 5, fig.width = 5}
ggplot(apajoin, aes(x = chla_n, y = Pg)) + 
  geom_point()
```

## Exercise 3

Repeat the above examples using the Sapelo Island dataset.

1. Import the nutrient data for the `sapdcnut` station using `import_local` and `qaqc`. 
1. Add a date column to the nutrients data using `mutate` from dplyr and `date` from lubridate. 
1. Floor the date columns in the nutrients data and metabolism data (`sapeco` from exercise 2) using`unit = 'month'` with  `floor_date` from lubridate.
1. Group and summarize the nutrients and metabolism data by the floor dates using `group_by` and `summarise`.  For the summaries, take the average of the metabolism estimates and any of the nutrient parameters of your choosing. 
1. Combine the summarized datasets Use `inner_join()` to combine the datasets by the floored date column.
1. Make a scatterplot in ggplot by mapping the mean nutrients and mean metabolism summaries to the x, y aesthetics. 
